---
title: "Bayesian Hierarchical Modelling of Fish Weights"
date: last-modified
author: Richard Marek
lang: en
format:
  html:
    html-math-method: katex
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    toc-title: Contents
    toc: true
    toc-depth: 3
    toc-location: right
    toc-expand: true
    self-contained-math: true
    fig-cap: true
    number-sections: true
    page-layout: article
    theme: cosmo
    max-width: 900px
execute:
  echo: true
  warning: false
  message: false
  error: false
  include: true
  cache: true
  fig-cap: true
crossref:
   tbl-title: "**Table**"
   fig-title: "**Figure**"
   tbl-labels: arabic
   fig-labels: arabic
   title-delim: "**.**"
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| echo: false
library("tidyverse")
library("runjags")
library(ggplot2)
library(HDInterval)
library("bayestestR")
library(TailRank)
library(Biobase)
library(invgamma)
library(rjags)
library(DT)
library(knitr)
library(kableExtra)
library(scales)

dataset <- read.csv(file = "Fish.csv", header = TRUE)
```

## Introduction

Hierarchical Bayesian models are widely used in actuarial science,
insurance, and catastrophe modelling, where data are naturally grouped —
for example, losses by peril type, geographic region, or line of
business. The central advantage of a hierarchical (multilevel) model is
**partial pooling**: groups with sparse data borrow strength from the
overall population, while groups with abundant data are primarily driven
by their own observations. This adaptive shrinkage is especially
valuable when group sample sizes are highly unequal, which is typical in
practice.

This project demonstrates these principles on the publicly available
[Fish Market
dataset](https://www.kaggle.com/datasets/vipullrathod/fish-market){target="_blank"},
fitting a Bayesian hierarchical Normal model to fish weight data across
seven species using JAGS (Just Another Gibbs Sampler). The methodology —
semi-conjugate priors, Half-Cauchy hyperpriors, and MCMC simulation —
transfers directly to grouped loss modelling contexts in (re)insurance.

## Data Description

The **Fish Market** dataset is freely available on
[Kaggle](https://www.kaggle.com/datasets/vipullrathod/fish-market){target="_blank"}.
It contains the following variables.

-   **Numeric variables**:
    -   `Height`, `Width`, `Length1`, `Length2`, `Length3` (dimensions
        in centimetres — height, width, and length measured in three
        different ways)
    -   `Weight` (weight in grams)
-   **Categorical variable**:
    -   `Species` — classifies fish into 7 categories:
        -   `Bream`
        -   `Parkki`
        -   `Perch`
        -   `Pike`
        -   `Roach`
        -   `Smelt`
        -   `Whitefish`

> **Note on Parkki:** No reliable photographic source for a fish called
> "Parkki" could be found. It is likely that this species name is
> misspelled in the dataset — it may refer to *Abramis bjoerkna* (White
> Bream / Silver Bream), known as "Pasuri" or "Blicca" in some European
> languages.

In this project we focus on comparing fish weight across species. The
dataset contains 159 observations in total, one of which was removed due
to a zero weight value.

```{r}
data <- dataset |>
  select(Species, y = Weight) |>
  filter(!is.na(y))

data[data$y == 0, ]
data <- data |> filter(y > 0)
```

## Exploratory Data Analysis

We start by plotting a boxplot of the raw data. It is immediately
apparent that individual species display substantially different weight
distributions. Median weights range from single grams (`Smelt`) to
approximately 600 g (`Bream`). Box widths — representing interquartile
ranges — differ markedly, reflecting heterogeneous within-species
variability. Sample sizes are also unbalanced: `Perch` has the most
observations, while `Whitefish` has the fewest. For `Perch` there is a
visible hint of two distinct sub-groups, possibly young versus adult
fish or two sub-species. Several outliers are present in `Roach`.

```{r}
data |>
  ggplot(aes(y = Species, x = y, fill = Species)) +
  geom_boxplot(color = "black", alpha = 0.3, outlier.shape = NA) +
  geom_jitter(pch = 21, size = 2, width = 0, height = 0.1) +
  labs(x = "Weight (g)", y = "Species")
```

The summary table below confirms the large cross-species differences.
Mean weight ranges from 11.2 g (`Smelt`) to 718.7 g (`Pike`), and sample
variances differ by several orders of magnitude — `Smelt` has a standard
deviation of 17.1 while `Pike` has a sample variance of 244,175.

The most represented species is `Perch` (56 fish); the least represented
is `Whitefish` (6 fish).

```{r}
ex_interactive <- data |>
  group_by(Species) |>
  summarise(
    `Observations` = n(),
    `Mean`         = mean(y),
    `Sample Variance` = var(y),
    .groups = "drop"
  ) |>
  mutate(
    `Observations`    = number(`Observations`, big.mark = ",", accuracy = 1),
    `Mean`            = number(`Mean`, big.mark = ",", accuracy = 0.1),
    `Sample Variance` = number(`Sample Variance`, big.mark = ",", accuracy = 0.1)
  )

DT::datatable(
  ex_interactive,
  rownames = FALSE,
  extensions = 'Buttons',
  options = list(
    dom = 'Bt',
    buttons = c('copy', 'csv', 'excel'),
    pageLength = 10
  )
) |>
  DT::formatStyle(
    columns = names(ex_interactive),
    fontSize = '14px',
    `text-align` = 'right'
  )
```

We intend to work with a Normal likelihood model, so we first assess
normality via the Shapiro–Wilk test.

```{r}
normality_tests_original <- data |>
  group_by(Species) |>
  summarise(
    `Observations` = n(),
    `p-value` = ifelse(shapiro.test(y)$p.value < 0.00001,
                       "< 0.00001",
                       format(round(shapiro.test(y)$p.value, 5), scientific = FALSE)),
    "H\u2080: data are normally distributed" = ifelse(
      shapiro.test(y)$p.value < 0.05,
      "Rejected in favour of H\u2081",
      "Not rejected"
    )
  ) |>
  arrange(`p-value`)

datatable(
  normality_tests_original,
  rownames = FALSE,
  options = list(
    dom = 'Bt',
    pageLength = 7,
    buttons = c('copy', 'csv', 'excel')
  )
) |>
  formatStyle(
    "H\u2080: data are normally distributed",
    backgroundColor = styleEqual(
      c("Rejected in favour of H\u2081", "Not rejected"),
      c('#ffdddd', '#ddffdd')
    )
  )
```

At the 5% significance level we reject normality for `Perch`, `Pike`,
`Smelt`, and `Roach`.

Because four species fail the normality test, a transformation is
needed. We apply a log transformation.

```{r}
data_jags <- data |>
  mutate(
    log_y  = log(y),
    group  = as.numeric(as.factor(Species))
  )
```

Shapiro–Wilk test on log-transformed data:

```{r}
normality_tests_log <- data_jags |>
  group_by(Species) |>
  summarise(
    `Observations` = n(),
    `p-value` = ifelse(shapiro.test(log_y)$p.value < 0.00001,
                       "< 0.00001",
                       format(round(shapiro.test(log_y)$p.value, 5), scientific = FALSE)),
    "H\u2080: data are normally distributed" = ifelse(
      shapiro.test(log_y)$p.value < 0.05,
      "Rejected in favour of H\u2081",
      "Not rejected"
    )
  ) |>
  arrange(`p-value`)

datatable(
  normality_tests_log,
  rownames = FALSE,
  options = list(
    dom = 'Bt',
    pageLength = 7,
    buttons = c('copy', 'csv', 'excel')
  )
) |>
  formatStyle(
    "H\u2080: data are normally distributed",
    backgroundColor = styleEqual(
      c("Rejected in favour of H\u2081", "Not rejected"),
      c('#ffdddd', '#ddffdd')
    )
  )
```

After the log transformation, normality is rejected only for `Perch`,
which we keep in mind during interpretation.

However, during further development we found that the hierarchical model
was numerically unstable under the plain log transformation

$$Y_j = \ln(X_j), \quad j = 1, \dots, 7,$$

where the index $j$ denotes the species. We therefore searched for a
shifted log transformation of the form

$$Y_j = \ln(X_j + c), \quad j = 1, \dots, 7,$$

where $c > 0$ is a constant to be chosen carefully.

Two competing requirements create a tension: a larger $c$ improves
numerical stability, but also tends to compress the data and reduce
p-values in the final inference, which can affect the interpretation of
results.

After several trials we settled on the compromise value $c = 50$, which
provides sufficient stability while keeping the effect on statistical
outputs acceptable. The final transformation is

$$Y_j = \ln(X_j + 50), \quad j = 1, \dots, 7.$$

```{r}
c_shift <- 50

data_jags <- data |>
  mutate(
    transf_y = log(y + c_shift),
    group    = as.numeric(as.factor(Species))
  )
```

Shapiro–Wilk test on the shifted log-transformed data:

```{r}
normality_tests_transf <- data_jags |>
  group_by(Species) |>
  summarise(
    `Observations` = n(),
    `p-value` = ifelse(shapiro.test(transf_y)$p.value < 0.00001,
                       "< 0.00001",
                       format(round(shapiro.test(transf_y)$p.value, 5), scientific = FALSE)),
    "H\u2080: data are normally distributed" = ifelse(
      shapiro.test(transf_y)$p.value < 0.05,
      "Rejected in favour of H\u2081",
      "Not rejected"
    )
  ) |>
  arrange(`p-value`)

datatable(
  normality_tests_transf,
  rownames = FALSE,
  options = list(
    dom = 'Bt',
    pageLength = 7,
    buttons = c('copy', 'csv', 'excel')
  )
) |>
  formatStyle(
    "H\u2080: data are normally distributed",
    backgroundColor = styleEqual(
      c("Rejected in favour of H\u2081", "Not rejected"),
      c('#ffdddd', '#ddffdd')
    )
  )
```

At the 5% level, normality is still rejected for `Perch` and `Smelt`,
which we keep in mind during interpretation.

A boxplot of the transformed data confirms that even after
transformation the species differ substantially. Medians fall roughly
between 4 and 7 on the transformed scale, and box widths still vary
considerably. Notable outliers remain — a very low value for `Perch`,
and two extreme observations each for `Smelt` and `Roach`.

```{r}
data_jags |>
  ggplot(aes(y = Species, x = transf_y, fill = Species)) +
  geom_boxplot(color = "black", alpha = 0.3, outlier.shape = NA) +
  geom_jitter(pch = 21, size = 2, width = 0, height = 0.1) +
  labs(x = "Transformed weight", y = "Species")
```

Histograms of the transformed data suggest approximately Normal
within-species distributions for most species. `Perch` still shows a
bimodal hint. `Parkki` appears roughly uniform, and `Whitefish` has too
few observations (6) to draw reliable conclusions.

```{r}
histogram <- data_jags |>
  ggplot(aes(x = transf_y, fill = Species)) +
  geom_histogram(bins = 30, color = "black", alpha = 0.7) +
  facet_wrap(~Species, scales = "free") +
  labs(x = "Transformed weight", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")
print(histogram)
```

QQ-plots reveal light tails for `Perch`, heavy tails for `Roach` (more
extreme values than a Normal would predict), a slight left tail for
`Pike`, and two exceptionally high observations for `Smelt` relative to
Normal quantiles.

```{r}
ggplot(data_jags, aes(sample = transf_y)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  facet_wrap(~Species, scales = "free") +
  labs(x = "Theoretical quantiles", y = "Sample quantiles") +
  theme_minimal()
```

```{r}
mapping_table <- data_jags |> select(Species, group) |> distinct()
```

## Model

We consider a hierarchical model for the transformed data
$Y_{j} = \ln(X_{j} + 50)$, built on the standard Normal model with
unknown mean and variance.

For each species $j = 1, \dots, 7$ we observe i.i.d. Gaussian random
variables
$(Y_{j,1}, \dots, Y_{j,n_j}) \sim \text{i.i.d. } \mathcal{N}(\mu_j, \sigma_j^2)$.

A sufficient statistic for $(\mu_j, \sigma_j^2)$ is

$$
\left( \sum_{i=1}^{n_j} Y_{j,i},\ \sum_{i=1}^{n_j} Y_{j,i}^2 \right),
$$

equivalently the sample mean and sample variance $(\bar{Y}_j, S^2_j)$.

The joint density expressed in terms of the sufficient statistics is

$$
f(\bar{y}_j, s^2_j \mid \mu_j, \sigma_j^2) = (2\pi\sigma_j^2)^{-\frac{n_j}{2}} \exp\left[-\frac{(n_j-1)s_j^2}{2\sigma_j^2}\right] \exp\left[-\frac{n_j}{2\sigma_j^2}(\bar{y}_j - \mu_j)^2\right].
$$

The likelihood function is

$$
L(\mu_j, \sigma_j^2 \mid \bar{y}_j, s_j^2) = \exp \left[ - \frac{(n_j-1)s_j^2}{2\sigma_j^2}\right] \exp\left[-\frac{n_j}{2\sigma_j^2}(\bar{y}_j - \mu_j)^2\right].
$$

**Semi-conjugate prior distributions:**

Variance (equivalently, precision): $$
\sigma^2_j \sim \text{InvGamma}\left(\frac{b_0}{2}, \frac{b_0\sigma_H^2}{2} \right), \quad \text{equivalently } \frac{1}{\sigma_j^2} \sim \text{Gamma}\left(\frac{b_0}{2}, \frac{b_0\sigma_H^2}{2} \right).
$$

Mean conditional on variance: $$
\mu_j \mid \sigma_j^2 \sim \mathcal{N}\left(\mu_H, \frac{\sigma_j^2}{a_0}\right).
$$

Joint prior density: $$
p(\mu_j, \sigma_j^2) = p(\sigma_j^2)\, p(\mu_j \mid \sigma_j^2).
$$

**Posterior distributions:**

Variance (equivalently, precision): $$
\sigma_j^2 \mid \bar{y}_j, s_j^2 \sim \text{InvGamma}\left( \frac{b_{j,n}}{2}, \frac{b_{j,n}\sigma_{j,n}^2}{2} \right).
$$

Mean conditional on variance: $$
\mu_j \mid \sigma_j^2, \bar{y}, s^2 \sim \mathcal{N} \left(\mu_{j,n}, \frac{\sigma^2}{a_{j,n}}\right).
$$

Joint posterior density: $$
p(\mu_j, \sigma_j^2 \mid \bar{y}_j, s_j^2) = p(\sigma_j^2 \mid \bar{y}_j, s_j^2)\, p(\mu_j \mid \sigma_j^2, \bar{y}_j, s_j^2).
$$

**Posterior parameters:**

$$
a_{n,j} = a_{0}+n_j,
$$ $$
b_{n,j} = b_{0} + n_j,
$$ $$
\mu_{n,j} = \bar{y}_j \cdot \frac{n_j}{n_j + a_0} + \mu_H \cdot \frac{a_0}{n_j + a_0},
$$ $$
b_{n,j} \sigma_{n,j}^2 = (n_j - 1)s_j^2 + b_0 \sigma_H^2 + \frac{a_0 n_j}{a_0 + n_j} (\bar{y}_j - \mu_H)^2.
$$

**Parameter interpretation:**

-   $\mu_j$ — mean of observations $Y_{j,i}$
-   $\sigma_j^2$ — variance of observations $Y_{j,i}$
-   $1/\sigma_j^2$ — precision of observations $Y_{j,i}$
-   $\bar{y}_j$ — sample mean for species $j$
-   $s_j^2$ — sample variance for species $j$
-   $\mu_H$ — hyperprior mean of $\mu_j$
-   $\mu_{n,j}$ — posterior mean of $\mu_j \mid \bar{y}_j, s_j^2$
-   $1/\sigma_H^2 > 0$ — hyperprior mean of the precision $1/\sigma_j^2$
-   $1/\sigma_{n,j}^2 > 0$ — posterior mean of the precision
    $1/\sigma_j^2 \mid \bar{y}_j, s_j^2$
-   $a_0 > 0$ — size of the fictitious prior sample for $\mu$
-   $b_0 > 0$ — size of the fictitious prior sample for $\sigma^2$
-   $\frac{n_j}{n_j + a_0}$ — weight assigned to the data
-   $\frac{a_0}{n_j + a_0}$ — weight assigned to the prior

**Hyperparameters**

Since we cannot specify the hyperparameters $\mu_H$ and $\sigma_H$
directly, we introduce a hierarchical structure and treat them as random
variables with their own distributions.

For $\mu_H$ we choose a Normal distribution with a very large variance,
so as not to favour any particular value, while still not assigning
equal probability to arbitrarily extreme values:

$$
\mu_H \sim \mathcal{N}(5,\; 10^6), \quad \mu_H > 0.
$$

For the scale hyperparameter $\sigma_H$ we follow the recommendation in
[Gelman
(2006)](https://sites.stat.columbia.edu/gelman/research/published/taumain.pdf){target="_blank"}
and use a Half-Cauchy prior:

$$
\sigma_H \sim \text{Cauchy}(0, A), \quad \sigma_H > 0,
$$

where $A$ is a scale parameter controlling the variability of
$\sigma_H$. Larger values of $A$ allow heavier tails and greater
between-group heterogeneity.

Because JAGS does not support the Cauchy distribution directly, we use
the equivalence with a Student-$t$ distribution with one degree of
freedom:

$$
\text{Student-}t(0,\, A^{-2},\, 1) \;=\; \text{Cauchy}(0, A).
$$

Returning to the key expression for $\mu_{n,j}$ — since we now treat
$\mu_H$ and $\sigma_H$ as random rather than fixed, the posterior mean
becomes

$$
\mu_{n,j} = \bar{y}_j \cdot \frac{n_j}{n_j + a_0} + \mathbb{E}(\mu_H) \cdot \frac{a_0}{n_j + a_0}.
$$

This is a **shrinkage estimator**: a weighted average of the group's own
data mean and the global prior mean. The weights depend on sample size
relative to $a_0$.

-   Weight on data: $\dfrac{n_j}{n_j + a_0}$
-   Weight on prior: $\dfrac{a_0}{n_j + a_0}$

The more observations a group has, the more weight is placed on its own
data; with few observations the prior dominates. We set $a_0 = 6$ to
achieve a roughly balanced influence between data and prior, reflecting
the fact that species can differ substantially and data quality is not
always ideal.

`Whitefish` and `Perch` represent the two extremes in our dataset.
`Whitefish` has $n_7 = 6$ observations (the minimum), while `Perch` has
$n_3 = 56$ (the maximum).

For `Whitefish`: $$
\frac{n_7}{n_7 + a_0} = \frac{6}{6 + 6} = 0.5, \qquad \frac{a_0}{n_7 + a_0} = \frac{6}{6 + 6} = 0.5.
$$

For `Perch`: $$
\frac{n_3}{n_3 + a_0} = \frac{56}{56 + 6} \approx 0.9, \qquad \frac{a_0}{n_3 + a_0} = \frac{6}{56 + 6} \approx 0.1.
$$

Now we prepare the input data for JAGS (the output shows the chosen
hyperparameter values):

```{r}
jags_data <- list(
  y1 = data_jags$transf_y[data_jags$Species == "Bream"],
  y2 = data_jags$transf_y[data_jags$Species == "Parkki"],
  y3 = data_jags$transf_y[data_jags$Species == "Perch"],
  y4 = data_jags$transf_y[data_jags$Species == "Pike"],
  y5 = data_jags$transf_y[data_jags$Species == "Roach"],
  y6 = data_jags$transf_y[data_jags$Species == "Smelt"],
  y7 = data_jags$transf_y[data_jags$Species == "Whitefish"],
  N1 = sum(data_jags$Species == "Bream"),
  N2 = sum(data_jags$Species == "Parkki"),
  N3 = sum(data_jags$Species == "Perch"),
  N4 = sum(data_jags$Species == "Pike"),
  N5 = sum(data_jags$Species == "Roach"),
  N6 = sum(data_jags$Species == "Smelt"),
  N7 = sum(data_jags$Species == "Whitefish"),
  J  = 7,   # number of groups
  A  = 1,   # Half-Cauchy scale; larger A -> heavier tails for sigma_H
  a0 = 6,
  b0 = 6
)

cat(" A =", jags_data$A,  "\n",
    "a0 =", jags_data$a0, "\n",
    "b0 =", jags_data$b0, "\n")
```

The JAGS model specification:

```{r}
jags_model <- "
model {
  # Likelihoods — one block per species
  for (i in 1:N1) { y1[i] ~ dnorm(mu[1], tau[1]) }
  for (i in 1:N2) { y2[i] ~ dnorm(mu[2], tau[2]) }
  for (i in 1:N3) { y3[i] ~ dnorm(mu[3], tau[3]) }
  for (i in 1:N4) { y4[i] ~ dnorm(mu[4], tau[4]) }
  for (i in 1:N5) { y5[i] ~ dnorm(mu[5], tau[5]) }
  for (i in 1:N6) { y6[i] ~ dnorm(mu[6], tau[6]) }
  for (i in 1:N7) { y7[i] ~ dnorm(mu[7], tau[7]) }

  # Group-level priors
  for (j in 1:J) {
    mu[j]  ~ dnorm(muH, a0 * tau[j]) T(0,)
    tau[j] ~ dgamma(b0 / 2, (b0 / 2) * sigmaH^2) T(0,)
  }

  # Hyperpriors
  muH    ~ dnorm(5, 0.000001) T(0,)
  # Half-Cauchy via Student-t with 1 degree of freedom
  sigmaH ~ dt(0, pow(A, -2), 1) T(0,)   # pow(A, -2) = 1/A^2
}
"
cat(jags_model)
```

### Simulation

With the model specified, we simulate the posterior distributions of the
transformed variables $Y_j$.

```{r}
inits_list <- list(
  list(.RNG.name = "base::Wichmann-Hill",  .RNG.seed = 1),
  list(.RNG.name = "base::Marsaglia-Multicarry", .RNG.seed = 2),
  list(.RNG.name = "base::Super-Duper",    .RNG.seed = 3)
)
result <- run.jags(
  model    = jags_model,
  data     = jags_data,
  monitor  = c("mu", "tau", "muH", "sigmaH"),
  inits    = inits_list,          
  n.chains = 3,
  burnin   = 1000,
  sample   = 5000,
  adapt    = 1000,
  thin     = 1,
  summarise = TRUE,
  silent.jags = TRUE
)

MCMC <- result$mcmc |> as.matrix(iters = TRUE, chains = TRUE) |> as.data.frame()

mu_cols  <- paste0("mu[",  1:7, "]")
tau_cols <- paste0("tau[", 1:7, "]")
mu_mat   <- MCMC[, mu_cols]
tau_mat  <- MCMC[, tau_cols]
sigma2_mat <- 1 / tau_mat

# Back-transform to the original scale:
# Y_j ~ N(mu_j, sigma_j^2), Y_j = log(X_j + c_shift)
# => X_j = exp(Y_j) - c_shift = Z_j - c_shift, where Z_j ~ LogNormal(mu_j, sigma_j^2)
# E[X_j] = exp(mu_j + sigma_j^2 / 2) - c_shift
# Var[X_j] = (exp(sigma_j^2) - 1) * exp(2*mu_j + sigma_j^2)
expectation <- exp(mu_mat  + sigma2_mat / 2) - c_shift
variance    <- (exp(sigma2_mat) - 1) * exp(2 * mu_mat + sigma2_mat)
precision   <- 1 / variance

colnames(variance) <- paste0("var[", 1:ncol(variance), "]")

MCMC_orig <- bind_cols(
  as_tibble(expectation),
  as_tibble(variance)
)
```

Because we modelled the data on the transformed scale

$$
Y_j = \ln(X_j + 50); \quad Y_j \sim \mathcal{N}\!\left(\mu_{n,j},\, \sigma_{n,j}^{2}\right),
$$

interpreting results on the original weight scale requires a
back-transformation. Starting from

$$
e^{Y_j} = X_j + 50 \implies X_j = e^{Y_j} - 50,
$$

we substitute $Z_j = e^{Y_j}$, so

$$
X_j = Z_j - 50, \quad \text{where } Z_j \sim \operatorname{LogNormal}\!\left(\mu_{n,j},\, \sigma_{n,j}^{2}\right).
$$

Using standard lognormal moment formulas:

$$
\mathbb{E}[X_j] = \mathbb{E}[Z_j] - 50 = e^{\mu_{n,j} + \sigma_{n,j}^2/2} - 50,
$$

$$
\operatorname{Var}(X_j) = \operatorname{Var}(Z_j) = \left(e^{\sigma_{n,j}^2} - 1\right) \cdot e^{2\mu_{n,j} + \sigma_{n,j}^2}.
$$

For reference, the species–index mapping used throughout is:

```{r}
DT::datatable(
  mapping_table,
  rownames = FALSE,
  extensions = 'Buttons',
  options = list(
    dom = 'Bt',
    buttons = c('copy', 'csv', 'excel'),
    pageLength = 7,
    columnDefs = list(list(className = 'dt-center', targets = 0:1)),
    order = list(list(0, 'asc'))
  ),
  caption = "Species-to-index mapping"
) |>
  formatStyle(columns = names(mapping_table), fontSize = '14px')
```

Summary statistics for posterior parameters on the original scale (after
back-transformation). The table includes the MAP estimate, posterior
mean, standard deviation, and 95% credible interval.

```{r}
summary_table_interactive <- MCMC_orig |>
  pivot_longer(cols = everything(), names_to = "parameter", values_to = "value") |>
  mutate(
    parameter = case_when(
      str_detect(parameter, "mu\\[\\d+\\]")  ~ str_replace(parameter, "mu\\[(\\d+)\\]",  "\u03bc<sub>\\1</sub>"),
      str_detect(parameter, "var\\[\\d+\\]") ~ str_replace(parameter, "var\\[(\\d+)\\]", "\u03c3\u00b2<sub>\\1</sub>"),
      TRUE ~ parameter
    )
  ) |>
  group_by(parameter) |>
  summarise(
    `MAP`   = round(bayestestR::map_estimate(value)$MAP_Estimate, 1),
    `Mean`  = round(mean(value), 1),
    `SD`    = round(sd(value), 1),
    `2.5%`  = round(quantile(value, 0.025), 1),
    `97.5%` = round(quantile(value, 0.975), 1),
    .groups = "drop"
  )

DT::datatable(
  summary_table_interactive,
  rownames = FALSE,
  escape = FALSE,
  options = list(
    dom = 'Bt',
    pageLength = 15
  )
)
```

The table shows substantial cross-species differences in both posterior
means and variances. `Pike` (index 4) has both a high posterior mean
weight and a large variance. `Roach` (index 5) sits at a lower level
with less spread. `Whitefish` (index 7) shows an unusually high
posterior standard deviation for its variance parameter, likely due to
numerical instability compounded by the very small sample size (6
observations) and a weight distribution that appears closer to Uniform
than Normal. Notably, the MAP estimate for `Whitefish` falls outside its
own 95% credible interval, suggesting convergence difficulties or a
misspecified model structure for that group.

## Posterior Mean Analysis

We plot histograms of posterior group means on the original scale. A
small number of extreme simulated values are excluded from the plot (at
most a few dozen out of 15,000 per species) to avoid axis distortion;
this does not affect interpretation materially.

All histograms share a broadly similar shape, though they differ in
location and spread, reflecting the true between-species weight
differences.

```{r}
mu_data_exp <- expectation |>
  select(starts_with("mu[")) |>
  pivot_longer(cols = everything(), names_to = "parameter", values_to = "value") |>
  mutate(index = as.integer(gsub("mu\\[|\\]", "", parameter))) |>
  left_join(mapping_table, by = c("index" = "group"))

mu_data_exp |>
  filter(parameter %in% paste0("mu[", 1:7, "]")) |>
  mutate(value = ifelse(Species == "Whitefish" & value > 1100, NA, value)) |>
  mutate(value = ifelse(Species == "Smelt"     & value > 300,  NA, value)) |>
  mutate(value = ifelse(Species == "Parkki"    & value > 500,  NA, value)) |>
  mutate(value = ifelse(Species == "Pike"      & value > 1200, NA, value)) |>
  mutate(value = ifelse(Species == "Roach"     & value > 350,  NA, value)) |>
  mutate(value = ifelse(Species == "Perch"     & value > 700,  NA, value)) |>
  mutate(value = ifelse(Species == "Bream"     & value > 800,  NA, value)) |>
  ggplot(aes(x = value, fill = Species)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30, alpha = 0.7, color = "white") +
  scale_fill_manual(values = rainbow(7)) +
  facet_wrap(~Species, scales = "free") +
  labs(title = "Posterior distributions of group means (original scale)",
       x = "Mean weight (g)",
       y = "Probability density") +
  theme_minimal() +
  theme(legend.position = "none")
```

The overlaid kernel density estimates below (x-axis capped at 1,200 g)
confirm the species separation and show different degrees of posterior
uncertainty. `Perch` and `Bream` have more symmetric and concentrated
posteriors; `Whitefish` and `Pike` show heavier tails; `Smelt` and
`Roach` have flatter, more spread-out densities.

```{r}
mu_data_exp |>
  filter(parameter %in% paste0("mu[", 1:7, "]")) |>
  ggplot(aes(x = value, color = Species, fill = Species)) +
  geom_density(alpha = 0.4) +
  scale_color_manual(values = rainbow(7)) +
  scale_fill_manual(values = rainbow(7)) +
  labs(title = "Kernel density estimates of posterior group means (original scale)",
       x = "Mean weight (g)",
       y = "Probability density") +
  xlim(0, 1100) +
  theme_minimal()
```

The following chart compares empirical sample means with posterior means
(back-transformed from MCMC). The model **shrinks** the posterior means
toward one another relative to the raw sample means — a direct
consequence of partial pooling. We also observe a rank swap between
`Roach` and `Parkki`: the two species switch their relative ordering
when moving from sample to posterior means.

```{r}
# Posterior mean summary
mu_summary <- expectation |>
  select(starts_with("mu[")) |>
  summarise(across(everything(), mean)) |>
  pivot_longer(cols = everything(), names_to = "parameter", values_to = "mean") |>
  mutate(index = as.integer(str_extract(parameter, "\\d+")))

# Empirical means
sample_means <- data |>
  group_by(Species) |>
  summarise(sample = mean(y), .groups = "drop") |>
  mutate(index = row_number())

plot_data <- left_join(sample_means, mu_summary, by = "index")

plot_data_long <- plot_data |>
  pivot_longer(cols = c(sample, mean), names_to = "type", values_to = "value") |>
  mutate(type = factor(type, levels = c("sample", "mean")))

ggplot(plot_data_long, aes(x = type, y = value, group = Species, color = Species)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  labs(title = "Empirical means vs. posterior means",
       x = NULL, y = "Weight (g)") +
  theme_minimal()
```

The same shrinkage pattern is visible when comparing empirical means
with MAP estimates, and the `Roach`/`Parkki` rank swap persists.

```{r}
map_summary <- expectation |>
  select(starts_with("mu[")) |>
  summarise(across(everything(), ~ bayestestR::map_estimate(.x)$MAP)) |>
  pivot_longer(cols = everything(), names_to = "parameter", values_to = "MAP") |>
  mutate(index = as.integer(str_extract(parameter, "\\d+")))

plot_data_map <- left_join(sample_means, map_summary, by = "index")

plot_data_map_long <- plot_data_map |>
  pivot_longer(cols = c(sample, MAP), names_to = "type", values_to = "value") |>
  mutate(type = factor(type, levels = c("sample", "MAP")))

ggplot(plot_data_map_long, aes(x = type, y = value, group = Species, color = Species)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  labs(title = "Empirical means vs. MAP estimates",
       x = NULL, y = "Weight (g)") +
  theme_minimal()
```

## Posterior Precision Analysis

Histograms of posterior group precisions on the original scale. As with
the means, the x-axis is trimmed to remove a small number of extreme
values.

The distributions display varying degrees of right-skewness. `Bream` is
closest to symmetric; `Smelt` and `Whitefish` have the heaviest right
tails. Absolute precision values are low (order $10^{-5}$ to $10^{-4}$),
consistent with the large weight variances observed in the raw data.

```{r}
tau_data_exp <- precision |>
  select(starts_with("tau[")) |>
  pivot_longer(cols = everything(), names_to = "parameter", values_to = "value") |>
  mutate(index = as.integer(gsub("tau\\[|\\]", "", parameter))) |>
  left_join(mapping_table, by = c("index" = "group"))

tau_data_exp |>
  mutate(value = ifelse(Species == "Whitefish" & value > 5e-05, NA, value)) |>
  mutate(value = ifelse(Species == "Smelt"     & value > 3e-04, NA, value)) |>
  mutate(value = ifelse(Species == "Parkki"    & value > 2e-04, NA, value)) |>
  mutate(value = ifelse(Species == "Roach"     & value > 2e-04, NA, value)) |>
  mutate(value = ifelse(Species == "Pike"      & value > 15e-06, NA, value)) |>
  filter(parameter %in% paste0("tau[", 1:7, "]")) |>
  ggplot(aes(x = value, fill = Species)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30, alpha = 0.7, color = "white") +
  scale_fill_manual(values = rainbow(7)) +
  facet_wrap(~Species, scales = "free") +
  labs(title = "Posterior distributions of group precisions (original scale)",
       x = "Precision",
       y = "Probability density") +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(size = 6))
```

The overlaid kernel density estimates (x-axis capped at 0.00015) show
that species differ not only in the location of their precision
distributions but also in their spread — some groups have considerably
more uncertainty in the precision estimate.

```{r}
tau_data_exp |>
  filter(parameter %in% paste0("tau[", 1:7, "]")) |>
  ggplot(aes(x = value, color = Species, fill = Species)) +
  geom_density(alpha = 0.2) +
  scale_color_manual(values = rainbow(7)) +
  scale_fill_manual(values = rainbow(7)) +
  labs(title = "Kernel density estimates of posterior group precisions (original scale)",
       x = "Precision",
       y = "Probability density") +
  xlim(0, 0.00015) +
  theme_minimal()
```

## Posterior Joint Distribution Analysis

We plot contours of the joint posterior distributions
$p(\mu_j, \tau_j \mid \mathbf{y})$ for each species.

For `Bream` the contours resemble a bivariate Normal ellipse, indicating
a relatively well-behaved joint posterior.

For `Parkki`, `Perch`, and `Roach` the contours are elongated in the
direction of the line $y \approx 1 - x$, indicating **negative posterior
dependence** between the mean and precision: higher mean values
correspond to lower precision estimates.

`Pike` displays a similar pattern but with the contours oriented more
vertically. `Whitefish` is the most extreme case — nearly vertical
contours suggest that precision varies widely while the mean stays
approximately fixed.

The contours for `Smelt` are strongly curved (approximately hyperbolic),
suggesting a complex mean–precision relationship or possible numerical
instability.

```{r}
plot_joint <- function(mu_data, tau_data, label) {
  comb <- bind_cols(
    mu_data  |> select(value) |> rename(mu = value),
    tau_data |> select(value) |> rename(tau = value)
  )
  ggplot(comb, aes(x = mu, y = tau)) +
    geom_point(alpha = 0.3, size = 0.5) +
    geom_density_2d_filled(alpha = 0.6, contour_var = "density", bins = 15) +
    geom_density_2d(color = "black", linewidth = 0.5, bins = 15) +
    scale_fill_manual(values = c("white", viridis::viridis(14, option = "C"))) +
    labs(x = "Mean", y = "Precision",
         title = paste0("Joint posterior contours — ", label)) +
    theme_minimal()
}

plot_joint(mu_data_exp |> filter(index == 1), tau_data_exp |> filter(index == 1), "Bream")
```

```{r}
plot_joint(mu_data_exp |> filter(index == 2), tau_data_exp |> filter(index == 2), "Parkki")
```

```{r}
plot_joint(mu_data_exp |> filter(index == 3), tau_data_exp |> filter(index == 3), "Perch")
```

```{r}
plot_joint(mu_data_exp |> filter(index == 4), tau_data_exp |> filter(index == 4), "Pike")
```

```{r}
plot_joint(mu_data_exp |> filter(index == 5), tau_data_exp |> filter(index == 5), "Roach")
```

```{r}
plot_joint(mu_data_exp |> filter(index == 6), tau_data_exp |> filter(index == 6), "Smelt")
```

Although the `Smelt` contour plot may suggest that $\tau_6$ reached
zero, a direct check confirms this did not occur — all 15,000 posterior
samples have $\tau_6 > 0$.

```{r}
data6tau <- tau_data_exp |> filter(index == 6) |> select(value) |> rename(tau = value)
sum(data6tau == 0)
```

```{r}
plot_joint(mu_data_exp |> filter(index == 7), tau_data_exp |> filter(index == 7), "Whitefish")
```

## Prior Distribution Analysis

We simulate the prior distributions of $Y_j$ by running JAGS with no
observed data and no likelihood — sampling from the prior alone.

```{r}
jags_data_prior <- list(
  J  = 7,
  A  = 1,
  a0 = 6,
  b0 = 6
)
cat(" A =", jags_data_prior$A,  "\n",
    "a0 =", jags_data_prior$a0, "\n",
    "b0 =", jags_data_prior$b0, "\n")
```

```{r}
jags_model_prior <- "
model {
  # Group-level priors only (no likelihood)
  for (j in 1:J) {
    mu[j]  ~ dnorm(muH, a0 * tau[j]) T(0,)
    tau[j] ~ dgamma(b0 / 2, (b0 / 2) * sigmaH^2) T(0,)
  }

  # Hyperpriors
  muH    ~ dnorm(5, 0.000001) T(0,)
  sigmaH ~ dt(0, pow(A, -2), 1) T(0,)
}
"
cat(jags_model_prior)
```

```{r}
result_prior <- run.jags(
  model    = jags_model_prior,
  data     = jags_data_prior,
  monitor  = c("mu", "tau", "muH", "sigmaH"),
  inits    = inits_list,        
  n.chains = 3,
  burnin   = 1000,
  sample   = 5000,
  adapt    = 1000,
  thin     = 1,
  summarise = TRUE,
  silent.jags = TRUE
)

MCMC_prior <- result_prior$mcmc |> as.matrix(iters = TRUE, chains = TRUE) |> as.data.frame()

mu_cols.prior  <- paste0("mu[",  1:7, "]")
tau_cols.prior <- paste0("tau[", 1:7, "]")
mu_mat.prior   <- MCMC_prior[, mu_cols.prior]
tau_mat.prior  <- MCMC_prior[, tau_cols.prior]
sigma2_mat.prior <- 1 / tau_mat.prior
```

> **Note on back-transformation of priors:** We do not back-transform
> prior samples to the original scale. The priors are intentionally
> weakly informative on the transformed scale; back-transforming would
> produce extreme values (up to infinity in some samples), which is an
> artefact of the transformation rather than a meaningful prior
> statement. All prior analysis therefore remains on the transformed
> scale.

Histograms of prior group means confirm that the distributions are
indeed weakly informative and broadly similar across species, as
intended.

```{r}
mu_data_prior <- mu_mat.prior |>
  pivot_longer(cols = everything(), names_to = "parameter", values_to = "value") |>
  filter(parameter %in% paste0("mu[", 1:7, "]")) |>
  mutate(index = as.integer(gsub("mu\\[|\\]", "", parameter))) |>
  left_join(mapping_table, by = c("index" = "group"))

mu_data_prior |>
  ggplot(aes(x = value, fill = Species)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30, color = "white") +
  scale_fill_hue(c = 60, l = 70) +
  facet_wrap(~Species, scales = "free") +
  xlim(0, 3500) +
  labs(title = "Prior distributions of group means (transformed scale)",
       x = "Mean of transformed weight",
       y = "Density") +
  theme_minimal() +
  theme(legend.position = "none")
```

Prior group precisions show a heavy right tail. Without axis truncation,
values up to $10^7$ appear — a consequence of the Gamma prior on
precision combined with the Half-Cauchy on $\sigma_H$. Most prior mass
is concentrated near zero, but the distribution is deliberately
fat-tailed to allow for potentially extreme precision values.

```{r}
tau_data_prior <- tau_mat.prior |>
  pivot_longer(cols = everything(), names_to = "parameter", values_to = "value") |>
  filter(parameter %in% paste0("tau[", 1:7, "]")) |>
  mutate(index = as.integer(gsub("tau\\[|\\]", "", parameter))) |>
  left_join(mapping_table, by = c("index" = "group"))

tau_data_prior |>
  ggplot(aes(x = value, fill = Species)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30, color = "white") +
  scale_fill_hue(c = 60, l = 70) +
  facet_wrap(~Species, scales = "free") +
  xlim(0, 10) +
  labs(title = "Prior distributions of group precisions (transformed scale)",
       x = "Precision (transformed scale)",
       y = "Density") +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(size = 6))
```

## Hyperparameter Analysis

Finally, we examine the marginal hyperparameter distributions by running
JAGS with no data and no likelihood or group-level priors — sampling
from the hyperpriors alone.

> **Note on hyperparameter priors:** We do not back-transform hyperparameter
> priors to the original scale. The priors are defined to be weakly informative
> on the transformed scale, where they are well-behaved and interpretable in
> terms of model regularisation. Back-transforming them would lead to extreme
> and unstable values, especially for scale-related parameters, which are not
> meaningful for prior interpretation. Therefore, all prior analysis and
> specification is kept on the transformed scale.

```{r}
jags_data_hyper <- list(A = 1)

jags_model_hyper <- "
model {
  muH    ~ dnorm(5, 0.000001) T(0,)
  sigmaH ~ dt(0, pow(A, -2), 1) T(0,)
}
"
```

```{r}
result_hyper <- run.jags(
  model    = jags_model_hyper,
  data     = jags_data_hyper,
  monitor  = c("muH", "sigmaH"),
  n.chains = 3,
  inits    = inits_list,         
  burnin   = 1000,
  sample   = 5000,
  adapt    = 1000,
  thin     = 1,
  summarise = TRUE,
  silent.jags = TRUE
)

MCMC_hyper <- result_hyper$mcmc |> as.matrix(iters = TRUE, chains = TRUE) |> as.data.frame()

mu_H    <- MCMC_hyper$muH
sigma_H <- MCMC_hyper$sigmaH
```

The histogram for $\mu_H$ confirms a very weakly informative, diffuse
distribution — consistent with our intent.

```{r}
ggplot(data.frame(mu_H = mu_H), aes(x = mu_H)) +
  geom_histogram(aes(y = after_stat(density)),
                 fill = "skyblue", color = "white", boundary = 0) +
  ggtitle(expression("Prior distribution of hyperparameter " * mu[H])) +
  labs(x = expression(mu[H]), y = "Probability density") +
  theme_minimal() +
  theme(text = element_text(size = 14))
```

The distribution of $\sigma_H$ has a pronounced right tail — typical of
the Half-Cauchy. Before axis truncation, values reach up to 200,000. We
choose the Half-Cauchy precisely because it is weakly informative yet
flexible: it concentrates most probability near zero (small
between-group heterogeneity) while still allowing extremely large values
(high heterogeneity), without hard-coding an upper bound. This is
advantageous when groups may vary dramatically — as they do here — but
we do not want to constrain that variation a priori.

```{r}
ggplot(data.frame(sigma_H = sigma_H), aes(x = sigma_H)) +
  geom_histogram(aes(y = after_stat(density)),
                 fill = "skyblue", color = "white") +
  ggtitle(expression("Prior distribution of hyperparameter " * sigma[H])) +
  labs(x = expression(sigma[H]), y = "Probability density") +
  xlim(0, 25) +
  theme_minimal() +
  theme(text = element_text(size = 14))
```

## Conclusion

We fitted a Bayesian hierarchical Normal model to log-transformed fish
weight data from the Fish Market dataset. Key findings and observations:

**Shrinkage.** The model pulls posterior group means toward the global
mean relative to the raw sample means. This effect is strongest for
species with few observations (`Whitefish`, $n = 6$) and weakest for the
most observed species (`Perch`, $n = 56$). The rank of `Roach` and
`Parkki` even swaps between the empirical and posterior estimates — an
example of how pooling can overturn naive comparisons when sample sizes
differ.

**Heterogeneity.** Despite the partial pooling, the seven species remain
clearly separated in their posterior mean distributions, confirming that
the dataset contains genuine between-species weight differences that the
hierarchical structure respects.

**Numerical stability.** The shifted log transformation
$Y = \ln(X + 50)$ was necessary to achieve MCMC convergence. The shift
constant $c = 50$ represents a deliberate compromise between stability
and minimal distortion of statistical inferences.

**Model limitations.** The normality assumption is violated for `Perch`
(bimodal distribution, likely a mixture of two sub-populations) and
`Smelt` (heavy-tailed outliers). Future extensions could replace the
Normal likelihood with a Student-$t$ likelihood for robustness, or fit a
finite mixture model for `Perch`. The `Whitefish` group — with only 6
observations — shows signs of convergence difficulties and warrants a
more informative prior or collection of additional data.

**Relevance to applied modelling.** The core technique — partial pooling
of group-level parameters within a hierarchical Bayesian framework — is
directly applicable to grouped loss modelling in insurance and
catastrophe modelling: for example, modelling claim frequencies or
severities pooled across geographic zones, peril types, or lines of
business, where data sparsity varies widely by group.
